Probabilities of polar alignment in circumbinary disks

### 2024-12-16 src/tex/ms.tex
Mention j in abstract

    Circumbinary gas disks that are misaligned to the binary orbital plane evolve toward either a coplanar or a polar-aligned configuration with respect to the binary host. The preferred alignment depends on the dynamics of the disk: whether it undergoes libration or nodal precession, with librating disks evolving to polar inclinations on the disk's viscous timescale. We quantify the fraction of binary star systems whose disks are expected to have polar orbits $f_\text{polar}$, finding that this quantity is heavily dependent on the host population's formation history. Modeling the disk as a geometrically thin ring, we use analytic and numerical methods to map sets of orbital parameters to dynamical states of the binary-disk system, calculating $f_\text{polar}$ by integrating over a large parameter space. Our findings can be used to make predictions given populations with well-understood distributions of binary eccentricity and initial mutual inclination. We also explore the consequences of the finite lifetime and non-zero radial extent of a real disk, which are incomplete evolution and a limit on the relative angular momentum imposed by disk breaking and gravitational instability, respectively.
\end{abstract}

@@

Through surveys of recent literature in circumbinary disk characterization \citep{czekala2019,zurlo2023}, \citet{ceppi2024} estimate that the fraction of circumbinary disks which are polar ($f_\text{polar}$) is $0.17 \pm 0.08$, though they acknowledge that the sample size is small ($\sim 15$ disks) and that the mutual inclination is hard to measure. They place this result in the context of a theoretical framework for a massless disk developed by \citet{farago2010,zanazzi2018} and conclude that observations are not telling the whole story. In this work, we use the results of \citet{martin2019} to extend this theoretical framework to massive disks.

Given enough time, a gas disk that is misaligned from one of the two stationary inclinations will bring itself to alignment through torques excited by warps that form as the disk undergoes differential precession \citep{bate2000,lubow2000}. Disks which are misaligned from coplanar only slightly (including the anti-aligned case where the mutual inclination $i\sim\pi$) undergo prograde or retrograde nodal precession \citep{papaloizou1995}. \citet{bate2000,lubow2000} show that this kind of motion will lead to coplanar alignment on the viscous timescale of the disk.

When the mutual inclination is $\sim\pi/2$ (near polar), the disk can librate about the eccentricity vector of the binary \citep[e.g.,][]{verrier2009,farago2010,doolin2011}. In sufficiently massive disks, a third kind of motion arises at high inclinations. Prograde precession takes over from retrograde precession near the anti-aligned point, and {\it crescent} orbits take over at inclinations between libration and high-inclination precession (\citealt{chen2019}, see also \citealt{zanazzi2018}), named because of the shape they form on a phase diagram. \citet{abod2022} showed that both librating and crescent orbiting disks decay to the stationary polar inclination \citep{martin2019} on the viscous timescale.

Thus, when the disk lifetime is longer than the viscous timescale \citep[which itself is similar to the precession timescale][]{bate2000}, it is appropriate to map the secular dynamics (which we call the dynamical state, e.g. nodal precession, libration or crescent orbits) to an end state of the disk (coplanar or polar, respectively). More challenging is mapping the parameters of the disk (e.g., binary eccentricity, initial inclination, disk mass, etc.) to a dynamical state. In general, this can be done using expensive 3D disk simulations (e.g., those shown in \citealt{martin2019}, \citealt{smallwood2019}, \& \citealt{abod2022}. It is not feasible to perform such simulations over a large parameter space, so we instead approximate a disk as a geometrically thin ring with angular momentum equivalent to that of the extended disk it represents. The literature provides a robust theoretical framework for the dynamics of such a ring, with analytical solutions \citep{farago2010,zanazzi2018,martin2019} as well as numerical $n$-body simulations \citep{abod2022,lepp2022} used to translate initial conditions into the ring's dynamical state.

>>>>

    Circumbinary gas disks that are misaligned to the binary orbital plane evolve toward either a coplanar or a polar-aligned configuration with respect to the binary host. The preferred alignment depends on the dynamics of the disk: whether it undergoes libration or nodal precession, with librating disks evolving to polar inclinations on the disk's viscous timescale. We quantify the fraction of binary star systems whose disks are expected to have polar orbits $f_\text{polar}$, extending previous work to include disks with non-zero mass. Our results suggest that massive disks are approximately as likely to become polar as disks in the massless limit, but dependence of $f_\text{polar}$ on the binary eccentricity can be significantly weakened. We also explore the consequences of the finite lifetime and non-zero radial extent of a real disk, which are incomplete evolution and a limit on the relative angular momentum imposed by disk breaking and gravitational instability, respectively. Our findings can be used to make predictions given populations with well-understood distributions of binary eccentricity, initial mutual inclination, and disk angular momentum.
\end{abstract}

@@

Through surveys of recent literature in circumbinary disk characterization \citep{czekala2019,zurlo2023}, \citet{ceppi2024} estimate that the fraction of circumbinary disks which are polar ($f_\text{polar}$) is $0.17 \pm 0.08$, though they acknowledge that the sample size is small ($\sim 15$ disks) and that the mutual inclination is hard to measure. They place this result in the context of a theoretical framework for a massless disk developed by \citet{farago2010} and \citet{zanazzi2018} to conclude that observations are not telling the whole story. In this work, we use the results of \citet{martin2019} to extend this theoretical framework to massive disks.

Given enough time, a gas disk that is misaligned from one of the two stationary inclinations will bring itself to alignment through torques excited by warps that form as the disk undergoes differential precession \citep{bate2000,lubow2000}. Disks which are misaligned from coplanar only slightly (including the anti-aligned case where the mutual inclination $i\sim\pi$) undergo prograde or retrograde nodal precession \citep{papaloizou1995}. \citet{bate2000} and \citet{lubow2000} show that this kind of motion will lead to coplanar alignment on the viscous timescale of the disk.

When the mutual inclination is $\sim\pi/2$ (near polar), the disk can librate about the eccentricity vector of the binary \citep[e.g.,][]{verrier2009,farago2010,doolin2011}. In sufficiently massive disks, a third kind of motion arises at high inclinations. Prograde precession takes over from retrograde precession near the anti-aligned point, and {\it crescent} orbits take over at inclinations between libration and high-inclination precession (\citealt{chen2019}, see also \citealt{zanazzi2018}), named because of the shape they form on a phase diagram. \citet{abod2022} showed that both librating and crescent orbiting disks decay to the stationary polar inclination \citep[see][]{martin2019} on the viscous timescale.

Thus, when the disk lifetime is longer than the viscous timescale \citep[which itself is similar to the precession timescale, see][]{bate2000}, it is appropriate to map the secular dynamics (which we call the dynamical state, e.g. nodal precession, libration or crescent orbits) to an end state of the disk (coplanar or polar, respectively). More challenging is mapping the parameters of the disk (e.g., binary eccentricity, initial inclination, disk mass, etc.) to a dynamical state. In general, this can be done using expensive 3D disk simulations (e.g., those shown in \citealt{martin2019}, \citealt{smallwood2019}, \& \citealt{abod2022}. It is not feasible to perform such simulations over a large parameter space, so we instead approximate a disk as a geometrically thin ring with angular momentum equivalent to that of the extended disk it represents. The literature provides a robust theoretical framework for the dynamics of such a ring, with analytical solutions \citep{farago2010,zanazzi2018,martin2019} as well as numerical $n$-body simulations \citep{abod2022,lepp2022} used to translate initial conditions into the ring's dynamical state.
